Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-08-10 jso.qmd
Matérn simulation figures
added Matérn simulation figures in R and R^2 (Figure 1 and Figure 2)

An classical way to describe the smoothness of a function is to identify
how many continuous derivatives of the function exist. This can be
characterized by Sobolev spaces [@adams2003sobolev].

@@

```{=tex}
\begin{defn}
The Matérn covariance function $K$ is defined by
\begin{equation}
K(u)=K_{\nu,\eta,\ell}(u):=\eta^2\frac{\left(\sqrt{2\nu}\frac{u}{\ell}\right)^{\nu}}{\Gamma(\nu)2^{\nu-1}}\mathcal{K}_{\nu}\left(\sqrt{2\nu}\frac{u}{\ell}\right)\ , u\geq0\label{eq:matern}
\end{equation}
where $\nu>0$ is the smoothness parameter, $\eta$ is the outputscale, $\ell$ is the lengthscale, and $\mathcal{K}_\nu$ is
the modified Bessel function \citep[10.25]{DLMF}. A multivariate extension $K(u)$, $u\in\mathbb{R}^d$ can be obtained by products of univariate covariance functions \eqref{eq:matern}.
\end{defn}
```
The Matérn covariance function can be expressed analytically when $\nu$
is a half-integer, the most popular values in the literature being
$\frac{1}{2}$, $\frac{3}{2}$ and $\frac{5}{2}$ [@rasmussen2006gaussian]. The parameter $\nu$,
called *smoothness parameter*, controls the decay of the covariance
function. As such, it is an appropriate measure of smoothness of a
random field. For example, @karvonen2023asymptotic showed that if a
function $f$ has a Sobolev index of $k$, then the smoothness parameter
estimate $\nu$ in \eqref{eq:matern} cannot be asymptotically less than
$k$. See the survey @porcu2024matern for additional results on the connection between the Matérn model and Sobolev spaces. An interesting result is that the asymptotic case
$\nu\rightarrow\infty$ coincides with the Gaussian kernel:
$K_\infty(u)=\exp(-u^{2}/2)$.

@@

It is expected that these two
values should be both high in the case of high squintability (fast
increase in $f$ early on), and both low in the case of low squintability
(any substantial increase in $f$ happens very late, close to the optimal
angle). This suggests that their product (@eq-squintability) should
provide a sensible measure of squintability. The multiplicative constant
$4$, which can be deemed arbitrary, does not change the interpretation
of the squintability metric $\varsigma$; it is here to adjust the range
of values of $\varsigma$ and simplify the explicit formula for
$\varsigma$ obtained later on.

@@

To compute the squintability metric (@eq-squintability) in practice,
several approaches are possible. The first one is to propose a
parametric model for $f$, and use it to obtain an explicit formula for
$\varsigma$. Numerical experiments suggest a scaled sigmoid shape as
described below. Define

@@

which depends on three additional parameters, $\theta_1$, $\theta_2$,
and $x_{\max}$, such that $f(0)=\theta_1$ and $f(x_{\max})=\theta_4$.
Under the parametric model (@eq-parametric), the squintability metric
(@eq-squintability) can be shown to be equal to

@@

Alternatively, one can estimate (@eq-squintability) in a nonparametric
way, for example by fitting $f$ using kernel regression, then
numerically estimate the angle at which $-f'$ attains its highest value.

@@

To compute squintability, 50 random bases are sampled and interpolated
to the optimal basis with a step size of 0.005. Index values and
projection distances are calculated for these interpolated bases and the
index values are averaged with a bin width of 0.005. A four-parameter
scaled logistic function is fitted to the index values against
projection distances, estimated by non-linear least squares. The
squintability measure is then calculated as
@eq-squintability-parametric.

@@

<!-- the best basis and it is calculated as the Frobenius norm in Equation -->

<!-- @eq-proj-dist. Plotting projection distances against index values for the  -->

@@

```{r}
#| label: fig-squintability
#| out.width: 100%
#| fig.cap: "Illustration of steps to calculate squintability. For a given projection pursuit problem defined by the shape to find, data dimension and the index function, 1) sample random bases given the orthonormality and projection distance contraint, 2) interpolate the sampled bases to the optimal basis and calculate the projection distance and the index value for each interpolated basis. 3) bin the index values by projection distances to obtain the average index value for each bin, 4) fit the scaled sigmoid function in Equation @eq-squintability to the binned index values against projection distances using non-linear least square, 5) calculate the squintability measure using Equation @eq-squintability-parametric with parameters estimated from the model."
# p1 <- sim_sine_6d_spline_projdist |> 
#   ggplot(aes(x = proj_dist, y = index_val)) +
#   geom_point(size = 0.1, alpha = 0.5) + 
#   labs(x = "Projection distance", y = "Index value") + 
#   theme_bw() + 
#   theme(axis.text = element_text(size = 20), axis.title = element_text(size = 30))
# 
# 
# dist_bin <- ceiling(sim_sine_6d_spline_projdist[["proj_dist"]] / 0.1) * 0.1
# dt <- sim_sine_6d_spline_projdist |>
#   dplyr::bind_cols(dist_bin = dist_bin) |>
#   dplyr::group_by(dist_bin) |>
#   dplyr::summarise(index_val = mean(index_val, na.rm = TRUE))
# 
# p2 <-  dt |> 
#   ggplot(aes(x = dist_bin, y = index_val)) + 
#   geom_line() +
#   geom_point() + 
#   labs(x = "Projection distance", y = "Index value") + 
#   theme_bw() + 
#   theme(axis.text = element_text(size = 20), axis.title = element_text(size = 30))
knitr::include_graphics(here::here("figures", "squintability.png"))
```

@@

The index properties, including smoothness and squintability, offer
numerical metrics to characterise the complexity of projection pursuit
optimisation problems. @tbl-smoothness-squintability presents the
parameters for calculating both metrics estimated from the Gaussian
process (variance, range, smooth, and nugget) and the scaled logistic
function ($\theta_1$ to $\theta_4$) for each case considered in the
simulation. The column "smooth" is used as the smoothness metric and the
column "squint" is calculated as equation @eq-squintability-parametric
as the squintability metric. @tbl-mod-output presents the results of
fitting a generalised linear model with a binomial family and a logit
link function to a sample of data in @tbl-mod-data, where all the three
components (success rate, JS hyper-parameters, and index properties) are
combined. The model suggests that JSO success rate is positively
associated with the two hyper-parameters, as well as with the index
properties: smoothness and squintability. Specifically, using 10 more
jellyfish and 10 more tries increases the odd ratio of success by
`r round(100 * (exp(broom::tidy(mod1)$estimate[6])-1), 2)`% and
`r round(100 * (exp(broom::tidy(mod1)$estimate[7])-1), 2)`%,
respectively. However, being flagged with long runtime and an increase
of data dimension reduce the success rate by
`r round(exp(broom::tidy(mod1)$estimate[5]) * 100, 2)`% and
`r round(exp(broom::tidy(mod1)$estimate[4]) * 100, 2)`%, respectively.
The variable `squintability` and `dimension` are significant, suggesting
their importance relative to JSO hyper-parameters in the optimisation
success. In defining a projection pursuit problem, factors such as the
shape-to-find, the index function used, and the data dimension,
determine the properties such as smoothness and squintability. These
metrics can then be compared across different problems to understand
their relative complexites. The results suggest that squintability has a
more significant influence than smoothness on the success rate of JSO.
Once the characteristics of the projection pursuit problem are fully
understood, increasing JSO hyper-parameters can enhance the search
effectiveness, however, it is also important to consider the resulting
increase in computational complexity.

>>>>

A classical way to describe the smoothness of a function is to identify
how many continuous derivatives of the function exist. This can be
characterized by Sobolev spaces [@adams2003sobolev].

@@

```{=tex}
\begin{defn}
The Matérn covariance function $K$ is defined by
\begin{equation}
K(u)=K_{\nu,\eta,\ell}(u):=\eta^2\frac{\left(\sqrt{2\nu}\frac{u}{\ell}\right)^{\nu}}{\Gamma(\nu)2^{\nu-1}}\mathcal{K}_{\nu}\left(\sqrt{2\nu}\frac{u}{\ell}\right)\ , u\geq0\label{eq:matern}
\end{equation}
where $\nu>0$ is the smoothness parameter, $\eta$ is the outputscale, $\ell$ is the lengthscale, and $\mathcal{K}_\nu$ is
the modified Bessel function \citep[10.25]{DLMF}. A multivariate extension $K(u)$, $u\in\mathbb{R}^d$ can be obtained by products of univariate covariance functions \eqref{eq:matern}.
\end{defn}
```
The Matérn covariance function can be expressed analytically when $\nu$
is a half-integer, the most popular values in the literature being
$\frac{1}{2}$, $\frac{3}{2}$ and $\frac{5}{2}$ [@rasmussen2006gaussian]. The parameter $\nu$,
called *smoothness parameter*, controls the decay of the covariance
function. As such, it is an appropriate measure of smoothness of a
random field, as shown by the simulations on @fig-matern-1d and @fig-matern-2d. For example, @karvonen2023asymptotic showed that if a
function $f$ has a Sobolev index of $k$, then the smoothness parameter
estimate $\nu$ in \eqref{eq:matern} cannot be asymptotically less than
$k$. See the survey @porcu2024matern for additional results on the connection between the Matérn model and Sobolev spaces. An interesting result is that the asymptotic case
$\nu\rightarrow\infty$ coincides with the Gaussian kernel:
$K_\infty(u)=\exp(-u^{2}/2)$.

```{r}
#| label: fig-matern-1d
#| out.width: 100%
#| fig.cap: |
#|   Five random simulations from a Gaussian Process defined on $\mathbb{R}$ with zero mean and Matérn-$\nu$ covariance function, with $\nu=1$ (left), $\nu=2$ (middle), and $\nu=4$ (right), showing that higher values of $\nu$ produce smoother curves.
knitr::include_graphics(here::here("figures", "matern_simulation_1d.png"))
```

```{r}
#| label: fig-matern-2d
#| out.width: 100%
#| fig.cap: |
#|   One random simulation from a Gaussian Process defined on $\mathbb{R}^2$ with zero mean and Matérn-$\nu$ covariance function, with $\nu=1$ (left), $\nu=2$ (middle), and $\nu=4$ (right), showing that higher values of $\nu$ produce smoother surfaces.
knitr::include_graphics(here::here("figures", "matern_simulation_2d.png"))
```

@@

It is expected that these two
values should be both high in the case of high squintability (fast
increase in $f$ early on), and both low in the case of low squintability
(any substantial increase in $f$ happens very late, close to the optimal
angle). This suggests that their product \eqref{eq-squintability} should
provide a sensible measure of squintability. The multiplicative constant
$4$, which can be deemed arbitrary, does not change the interpretation
of the squintability metric $\varsigma$; it is here to adjust the range
of values of $\varsigma$ and simplify the explicit formula for
$\varsigma$ obtained later on.

@@

To compute the squintability metric \eqref{eq-squintability} in practice,
several approaches are possible. The first one is to propose a
parametric model for $f$, and use it to obtain an explicit formula for
$\varsigma$. Numerical experiments suggest a scaled sigmoid shape as
described below. Define

@@

which depends on three additional parameters, $\theta_1$, $\theta_2$,
and $x_{\max}$, such that $f(0)=\theta_1$ and $f(x_{\max})=\theta_4$.
Under the parametric model (@eq-parametric), the squintability metric
\eqref{eq-squintability} can be shown to be equal to

@@

Alternatively, one can estimate \eqref{eq-squintability} in a nonparametric
way, for example by fitting $f$ using kernel regression, then
numerically estimate the angle at which $-f'$ attains its highest value.

@@

To compute squintability, 50 random bases are sampled and interpolated
to the optimal basis with a step size of 0.005. Index values and
projection distances are calculated for these interpolated bases and the
index values are averaged with a bin width of 0.005. A four-parameter
scaled logistic function is fitted to the index values against
projection distances, estimated by non-linear least squares. The
squintability measure is then calculated as
(@eq-squintability-parametric).

@@

<!-- the best basis and it is calculated as the Frobenius norm in equation -->

<!-- (@eq-proj-dist). Plotting projection distances against index values for the  -->

@@

```{r}
#| label: fig-squintability
#| out.width: 100%
#| fig.cap: |
#|   Illustration of steps to calculate squintability. For a given projection pursuit problem defined by the shape to find, data dimension and the index function, 1) sample random bases given the orthonormality and projection distance contraint, 2) interpolate the sampled bases to the optimal basis and calculate the projection distance and the index value for each interpolated basis. 3) bin the index values by projection distances to obtain the average index value for each bin, 4) fit the scaled sigmoid function in equation \eqref{eq-squintability} to the binned index values against projection distances using non-linear least square, 5) calculate the squintability measure using equation (@eq-squintability-parametric) with parameters estimated from the model.
# p1 <- sim_sine_6d_spline_projdist |> 
#   ggplot(aes(x = proj_dist, y = index_val)) +
#   geom_point(size = 0.1, alpha = 0.5) + 
#   labs(x = "Projection distance", y = "Index value") + 
#   theme_bw() + 
#   theme(axis.text = element_text(size = 20), axis.title = element_text(size = 30))
# 
# 
# dist_bin <- ceiling(sim_sine_6d_spline_projdist[["proj_dist"]] / 0.1) * 0.1
# dt <- sim_sine_6d_spline_projdist |>
#   dplyr::bind_cols(dist_bin = dist_bin) |>
#   dplyr::group_by(dist_bin) |>
#   dplyr::summarise(index_val = mean(index_val, na.rm = TRUE))
# 
# p2 <-  dt |> 
#   ggplot(aes(x = dist_bin, y = index_val)) + 
#   geom_line() +
#   geom_point() + 
#   labs(x = "Projection distance", y = "Index value") + 
#   theme_bw() + 
#   theme(axis.text = element_text(size = 20), axis.title = element_text(size = 30))
knitr::include_graphics(here::here("figures", "squintability.png"))
```

@@

The index properties, including smoothness and squintability, offer
numerical metrics to characterise the complexity of projection pursuit
optimisation problems. @tbl-smoothness-squintability presents the
parameters for calculating both metrics estimated from the Gaussian
process (variance, range, smooth, and nugget) and the scaled logistic
function ($\theta_1$ to $\theta_4$) for each case considered in the
simulation. The column "smooth" is used as the smoothness metric and the
column "squint" is calculated as equation (@eq-squintability-parametric)
as the squintability metric. @tbl-mod-output presents the results of
fitting a generalised linear model with a binomial family and a logit
link function to a sample of data in @tbl-mod-data, where all the three
components (success rate, JS hyper-parameters, and index properties) are
combined. The model suggests that JSO success rate is positively
associated with the two hyper-parameters, as well as with the index
properties: smoothness and squintability. Specifically, using 10 more
jellyfish and 10 more tries increases the odd ratio of success by
`r round(100 * (exp(broom::tidy(mod1)$estimate[6])-1), 2)`% and
`r round(100 * (exp(broom::tidy(mod1)$estimate[7])-1), 2)`%,
respectively. However, being flagged with long runtime and an increase
of data dimension reduce the success rate by
`r round(exp(broom::tidy(mod1)$estimate[5]) * 100, 2)`% and
`r round(exp(broom::tidy(mod1)$estimate[4]) * 100, 2)`%, respectively.
The variable `squintability` and `dimension` are significant, suggesting
their importance relative to JSO hyper-parameters in the optimisation
success. In defining a projection pursuit problem, factors such as the
shape-to-find, the index function used, and the data dimension,
determine the properties such as smoothness and squintability. These
metrics can then be compared across different problems to understand
their relative complexites. The results suggest that squintability has a
more significant influence than smoothness on the success rate of JSO.
Once the characteristics of the projection pursuit problem are fully
understood, increasing JSO hyper-parameters can enhance the search
effectiveness, however, it is also important to consider the resulting
increase in computational complexity.
